An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-09 cal_demo/ms.tex
Switched data and noise sections. Needs to be ironed out for new flow
touching up noise section after switching order with data section

@@

\section{Noise trends}\label{sec:noise}
Now we return to simulations to study the effect of noise in our system, and the consequences 
of an incomplete sky model. We run a suite of simulations varying the receiver noise and 
integration times, while forming antenna-pixel correlations, $\Cna$, for EPICal gain solutions, 
and simultaneously forming visibilities from the same E-field streams to find visibility-based 
gain solutions for comparison.

We use a simulated sky consisting of a 5~Jy calibrator source, and 49 other random sources 
with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$ 23 Jy). We generated 
these sources randomly, but kept them fixed for each run. As in section~\ref{sec:sim}, we use 
the MWA core layout (51 antennas), and a 4.4~m square tophat antenna voltage pattern. We 
simulate 64 frequency channels each of width 40~kHz. Because our calibration loop treats 
frequency channels independently, each channel can be treated as a separate trial of the 
simulation, and is used to better estimate the statistics.

To simulate receiver noise, we add a gaussian distributed complex random number to each 
antenna electric field measurement at each time stamp according to 
equation~\ref{eq:apply_gain}. The level of the receiver noise, $\sigma_r = \left<\left|
\widetilde{n}_a(f,t)\right|^2\right>$, is varied in different simulation runs. We include two limiting 
cases, where the receiver noise power is subdominant to the sky power ($\sigma_r = 
10.0$~Jy), and where the receiver dominates the noise ($\sigma_r=100.0$~Jy).

For each simulation run we find gain solutions using four methods:
\begin{enumerate}[i.]
\item EPICal, using a full sky model
\item EPICal, using a single point source model
\item Visibility-based, using a full sky model
\item Visibility-based, using a single point source model.
\end{enumerate}
In the cases where we use the full sky model, the model visibilities, $V^M_{ab}$ used in the 
calibration loop are created using all point sources in the simulated sky. For the single point 
source model, we only allow the calibration loop to be privy to the bright calibrator source. 
While we vary the model used for calibration, the simulated ``true sky'' remains fixed with all 50 
point sources.

For the EPICal calibration loops, we initialize our gain estimates with the true values 
(${g^{(0)}_a=g_a=1}$), and allow the estimate to be corrupted by the noise through ten 
iterations of the calibration loop. We adopt a damping factor $\damp=0.35$. Because EPICal 
updates the gain estimate at each calibration loop, \tcal, but retains a memory of previous 
iterations through the damping factor, the total integration time is not straightforward. We define 
an effective integration time by considering the relative weights of each previous $\Cna$ 
contributing to the current $g^{(n)}_a$ estimate, each with integration time \tcal. In the limit $n
\rightarrow \infty$, the geometric series converges to an effective integration time of
\begin{equation}
\teff = \tcal \times \frac{1+\damp}{1-\damp}.
\end{equation}
With the damping factor we adopted here, $\teff \approx 2.1 \times \tcal$.

We simultaneously form simulated visibilities by correlating all pairs of antenna E-field 
measurements (including the receiver noise). The E-fields are correlated for a duration equal to 
the effective integration time of the EPICal loop for comparison. We then find the visibility-
based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|V_{ab}-g_a g_b^* V^M_{ab}\right|^2,
\end{equation}
using both versions of the model visibilities described above.

 For each version of calibration, we observe the error in the gain estimates averaging over both 
 antennas and frequency channels.
\begin{equation}\label{eq:gain_error}
\sigma_g = \left[\frac{1}{N_f \Nant} \sum_f \sum_a \frac{\left|g^{(n)}_a(f)-g_a(f)\right|^2}{\left|g_a(f)\right|^2}\right]^{1/2}
\end{equation}

The results of our simulations are shown in figure~\ref{fig:errors}. The gain errors from our 
EPICal method are shown with solid lines, while the dashed lines represent the visibility-based 
calibration. The thick lines represent the results of using a full sky model, and the thin lines are 
from the single point source model.  In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise
in the sense that they integrate down with longer integrations with the expected $t^{-1/2}$ 
trend. 
The same level of gain error as visibility based solutions can be achieved with a longer integration time (or larger damping 
factor).

>>>>

\section{Noise trends}\label{sec:noise}
Now next study the effect of noise in our system, and the consequences 
of an incomplete sky model. We run a suite of simulations varying the receiver noise and 
integration times, while forming antenna-pixel correlations, $\Cna$, for EPICal gain solutions, 
and simultaneously forming visibilities from the same electric field measurement streams to find visibility-based 
gain solutions for comparison.

We use a simulated sky consisting of a 5~Jy calibrator source, and 49 other random sources 
with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$ 23 Jy). We generated 
these sources randomly, but kept them fixed for each run. 
We simulate the same antenna layout as in Section~\ref{sec:sim}. 
However, we increase the number of frequency channels simulated to 64.
Because our calibration loop treats 
frequency channels independently, each channel can be treated as a separate trial of the 
simulation, and is used to better estimate the statistics.

To simulate receiver noise, we add a gaussian distributed complex random number to each 
antenna electric field measurement at each time sample according to 
equation~\ref{eq:apply_gain}. The level of the receiver noise, $\sigma_r = \left<\left|
n_a(f,t)\right|^2\right>$, is varied in different simulation runs. We include two limiting 
cases, where the receiver noise power is subdominant to the sky power ($\sigma_r = 
10.0$~Jy), and where the receiver dominates the noise ($\sigma_r=100.0$~Jy).

For each simulation run we find gain solutions using four methods:
\begin{enumerate}[i.]
\item EPICal, using a full sky model
\item EPICal, using a single point source model
\item Visibility-based, using a full sky model
\item Visibility-based, using a single point source model.
\end{enumerate}
In the cases where we use the full sky model, the model visibilities, $V^M_{ab}$, used in the 
calibration loop are created using all point sources in the simulated sky. For the single point 
source model, we only allow the calibration loop to be privy to the bright calibrator source. 
While we vary the model used for calibration, the simulated ``true sky'' remains fixed with all 50 
point sources.

For the EPICal calibration loops, we initialize our gain estimates with the true values 
(${g^{(0)}_a=g_a=1}$), and allow the estimate to be corrupted by the noise through ten 
iterations of the calibration loop. We again adopt a damping factor $\damp=0.35$. Because EPICal 
updates the gain estimate at each calibration loop, \tcal, but retains a memory of previous 
iterations through the damping factor, the total integration time is not straightforward. We define 
an effective integration time by considering the relative weights of each previous $\Cna$ 
contributing to the current $g^{(n)}_a$ estimate, each with integration time \tcal. In the limit $n
\rightarrow \infty$, the geometric series converges to an effective integration time of
\begin{equation}
\teff = \tcal \times \frac{1+\damp}{1-\damp}.
\end{equation}
With the damping factor we adopted here, $\teff \approx 2.1 \times \tcal$.

We simultaneously form simulated visibilities, $V_{ab}$, by correlating all pairs of antenna electric field 
measurements (including the receiver noise). The electric field measurements are correlated for a duration equal to 
the effective integration time of the EPICal loop for comparison. We then find the visibility-
based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|V_{ab}-g_a g_b^* V^M_{ab}\right|^2,
\end{equation}
using both versions of the model visibilities described above.

 For each version of calibration, we observe the error in the gain estimates by averaging over both 
 antennas and frequency channels.
\begin{equation}\label{eq:gain_error}
\sigma_g = \left[\frac{1}{N_f \Nant} \sum_f \sum_a \frac{\left|g^{(n)}_a(f)-g_a(f)\right|^2}{\left|g_a(f)\right|^2}\right]^{1/2}
\end{equation}

The results of our simulations are shown in Figure~\ref{fig:errors}. The gain errors from our 
EPICal method are shown with solid lines, while the dashed lines represent the visibility-based 
calibration. The thick lines represent the results of using a full sky model, and the thin lines are 
from the single point source model.  In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise
in the sense that they integrate down with longer integrations with the expected $t^{-1/2}$ 
trend. 
The same level of gain error as visibility based solutions can be achieved with a longer integration time (or larger damping 
factor).

@@
